---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# forestploter

<!-- badges: start -->
[![R-CMD-check](https://github.com/adayim/forestploter/workflows/R-CMD-check/badge.svg)](https://github.com/adayim/forestploter/actions)
[![CRAN status](https://www.r-pkg.org/badges/version/forestploter)](https://CRAN.R-project.org/package=forestploter)
[![CRAN download](https://cranlogs.r-pkg.org/badges/grand-total/forestploter)](https://cran.r-project.org/package=forestploter)
[![Codecov test coverage](https://codecov.io/gh/adayim/forestploter/branch/main/graph/badge.svg)](https://app.codecov.io/gh/adayim/forestploter?branch=main)
<!-- badges: end -->

The goal of `forestploter` is to create publication-ready forest plots with minimal effort. This package offers more flexibility and customization options compared to other packages. The layout of the forest plot is determined by the provided dataset, and its elements are organized in a grid-like structure of rows and columns, much like a table. This design allows for easy manipulation of any plot component. For instance, the width of the confidence interval column can be controlled by adjusting the number of spaces in a blank character column.

## Installation

You can install the development version of `forestploter` from [GitHub](https://github.com/adayim/forestploter) with:

**Install from CRAN**
``` r
install.packages("forestploter")
```

**Install development version from GitHub**
```r
# install.packages("devtools")
devtools::install_github("adayim/forestploter")
```

## Basic Usage

The column names of the provided data will be used as the header of the plot. `forest()` draws the table and the confidence intervals. The rest is added with the pipe `|>`:

* `set_xaxis()` sets the limits, tick marks and scale of the x-axis, and adds vertical lines.
* `set_labs()` sets the title, x-axis labels, footnote, arrow labels and legend labels.
* `scale_sizes()` scales the point sizes by study weights.
* `forest_style()` and `set_style()` set the graphical parameters.

This is a basic example that demonstrates how to create a `forestplot`:

```{r example, dpi=300, fig.width  = 12, fig.height = 5.8, out.height = '40%'}
library(grid)
library(forestploter)

dt <- read.csv(system.file("extdata", "example_data.csv", package = "forestploter"))

# Indent the subgroup if there is a number in the placebo column
dt$Subgroup <- ifelse(is.na(dt$Placebo),
                      dt$Subgroup,
                      paste0("   ", dt$Subgroup))

# NA to blank
dt$Treatment <- ifelse(is.na(dt$Treatment), "", dt$Treatment)
dt$Placebo <- ifelse(is.na(dt$Placebo), "", dt$Placebo)
dt$se <- (log(dt$hi) - log(dt$est))/1.96

# Add a blank column for the forest plot to display CI.
# Adjust the column width with space.
dt$` ` <- paste(rep(" ", 20), collapse = " ")

# Create confidence interval column to display
dt$`HR (95% CI)` <- ifelse(is.na(dt$se), "",
                             sprintf("%.2f (%.2f to %.2f)",
                                     dt$est, dt$low, dt$hi))

# Define a style, it can be reused for other plots
st <- forest_style(base_size = 10,
                   arrow_type = "closed",
                   footnote = gpar(col = "blue", cex = 0.6))

p <- forest(dt[,c(1:3, 20:21)],
            est = dt$est,
            lower = dt$low,
            upper = dt$hi,
            sizes = dt$se,
            ci_column = 4,
            ref_line = 1,
            style = st) |>
  set_xaxis(xlim = c(0, 4), ticks_at = c(0.5, 1, 2, 3)) |>
  set_labs(arrow = c("Placebo Better", "Treatment Better"),
           footnote = "This is the demo data. Please feel free to change\nanything you want.")

# Print plot
plot(p)

```

The plot is a `gtable`, so it can be saved with `ggplot2::ggsave()` and combined with other plots, for example with `patchwork::wrap_elements()`. Set `fit` in `forest_style()` to let the plot fill the space it is given. The arguments of `forest()` used before version 1.2.0, such as `xlim` or `footnote`, and themes created with `forest_theme()` still work. The old arguments give a message pointing to `set_xaxis()` or `set_labs()`, and `?forest_theme` shows how the theme settings map onto `forest_style()`.

## Editing the Plot

You may want to change the color or font of certain columns, insert text into specific rows, or add an underline to separate groups. The `edit_plot`, `add_text`, `insert_text`, and `add_border` functions are designed for these purposes. They can be chained with the pipe as well, after `set_xaxis()`, `set_labs()`, `scale_sizes()` and `set_style()`, which build the plot again:

```{r edit, dpi=300, fig.width  = 12, fig.height = 6.3, out.height = '40%'}
g <- p |>
  # Edit text in row 3
  edit_plot(row = 3, gp = gpar(col = "red", fontface = "italic")) |>
  # Bold grouping text
  edit_plot(row = c(2, 5, 8, 11, 15, 18),
            gp = gpar(fontface = "bold")) |>
  # Insert text at the top
  insert_text(text = "Treatment group",
              col = 2:3,
              part = "header",
              gp = gpar(fontface = "bold")) |>
  # Add underline at the bottom of the header
  add_border(part = "header", row = 1, where = "top") |>
  add_border(part = "header", row = 2, where = "bottom") |>
  add_border(part = "header", row = 1, col = 2:3,
             gp = gpar(lwd = 2)) |>
  # Edit the background of row 5
  edit_plot(row = 5, which = "background",
            gp = gpar(fill = "darkolivegreen1")) |>
  # Insert text
  insert_text(text = "This is a long text. Age and gender summarised above.\nBMI is next",
              row = 10,
              just = "left",
              gp = gpar(cex = 0.6, col = "green", fontface = "italic")) |>
  add_border(row = 10, col = 1:3, where = "top")

plot(g)

```

Remember to add 1 to the row number if you have inserted any text before, as the row number will change after inserting text.

## Complex Usage

If you want to draw CIs in multiple columns, you only need to provide a vector of the column positions in the data. As shown in the example below, the CIs will be drawn in columns 3 and 5, with the first and second `est`, `lower`, and `upper` values corresponding to those columns.

For more complex scenarios, such as drawing CIs by groups, you can provide an additional set of `est`, `lower`, and `upper` values. If the number of `est`, `lower`, and `upper` sets is greater than the number of CI columns, the values will be reused. In the example below, `est_gp1` and `est_gp2` are drawn in columns 3 and 5 as **group 1**, while `est_gp3` and `est_gp4` are drawn in the same columns as **group 2**.

This is an example of multiple CI columns and groups:

```{r multiple,  dpi=300, fig.width  = 12, fig.height = 7.3, out.height = '80%'}

# Add a blank column for the second CI column
dt$`   ` <- paste(rep(" ", 20), collapse = " ")

p <- forest(dt[,c(1:2, 20, 3, 22)],
            est = list(dt$est_gp1,
                       dt$est_gp2,
                       dt$est_gp3,
                       dt$est_gp4),
            lower = list(dt$low_gp1,
                         dt$low_gp2,
                         dt$low_gp3,
                         dt$low_gp4),
            upper = list(dt$hi_gp1,
                         dt$hi_gp2,
                         dt$hi_gp3,
                         dt$hi_gp4),
            ci_column = c(3, 5),
            ref_line = 1,
            nudge_y = 0.2,
            style = forest_style(base_size = 10,
                                 footnote = gpar(col = "blue"))) |>
  set_xaxis(x_trans = "log") |>
  set_labs(arrow = c("Placebo Better", "Treatment Better"),
           legend_title = "GP",
           legend_labels = c("Trt 1", "Trt 2"))

plot(p)


```
