---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# forestploter

<!-- badges: start -->
<!-- badges: end -->

The goal of forestploter is to create a publication-ready forest plot with little effort. This package provide some extra displays compared to other packages.

## Installation

You can install the development version of forestploter from [GitHub](https://github.com/adayim/forestploter) with:

``` r
# install.packages("devtools")
devtools::install_github("adayim/forestploter")
```

## Basic usage

This is a basic example which shows you how to create a forestplot:

```{r example, dpi=300, fig.width  = 12, fig.height = 6, out.height = '30%'}
library(forestploter)


dt <- read.csv(system.file("extdata", "example_data.csv", package = "forestploter"))

# indent the subgroup if there is a number in the placebo column
dt$Subgroup <- ifelse(is.na(dt$Placebo), 
                      dt$Subgroup,
                      paste0("   ", dt$Subgroup))

# NA to blank
dt$Treatment <- ifelse(is.na(dt$Treatment), "", dt$Treatment)
dt$Placebo <- ifelse(is.na(dt$Placebo), "", dt$Placebo)
dt$se <- (log(dt$hi) - log(dt$est))/1.96

# Add blank column for the forest plot to display CI
dt$` ` <- paste(rep(" ", 5), collapse = " ")

# Create confidence interval column to display
dt$`HR (95% CI)` <- ifelse(is.na(dt$se), "",
                             sprintf("%.2f (%.2f to %.2f)",
                                     dt$est, dt$low, dt$hi))

# Define theme
tm <- forest_theme(base_size = 10,
                   refline_col = "red")

p <- forest(dt[,c(1:3, 20:21)],
            est = dt$est,
            lower = dt$low, 
            upper = dt$hi,
            sizes = dt$se,
            ci.column = 4,
            ref.line = 1,
            ci.column.width = 2,
            arrow.lab = c("Placebo Better", "Treatment Better"),
            tick.breaks = c(0.5, 1, 2, 4),
            theme = tm)

# Draw plot
plot(p)


```

## Complex usage

If you want to draw CI to multiple columns, only need to provide a vector of the position of the columns to be drawn in the data. As seen in the example below, the CI will be drawn in the column 3 and 5. The first and second `est`, `lower` and `upper` will be drawn in column 3 and column 5. 

For a more complex example, you may want to draw CI by groups. The solution is simple, just provide another set of `est`, `lower` and `upper`. If the number of provided `est`, `lower` and `upper` are larger than the column number to draw CI, then the `est`, `lower` and `upper` will be reused. As it is shown in the example below, `est_gp1` and `est_gp2` will be drawn in column 3 and column 5 as normal. But `est_gp3` and `est_gp4` hasn't been used, those will be drawn to column 3 and column 5 again. So, the `est_gp1` and `est_gp2` will be considered as group 1, `est_gp3` and `est_gp4` group 2.

This is an example of multiple CI columns and groups:

```{r multiple,  dpi=300, fig.width  = 14, fig.height = 7, out.height = '120%'}

# Add blank column for the second CI column
dt$`   ` <- paste(rep(" ", 5), collapse = " ")

p <- forest(dt[,c(1:2, 20, 3, 22)],
            est = list(dt$est_gp1,
                       dt$est_gp2,
                       dt$est_gp3,
                       dt$est_gp4),
            lower = list(dt$low_gp1,
                         dt$low_gp2,
                         dt$low_gp3,
                         dt$low_gp4), 
            upper = list(dt$hi_gp1,
                         dt$hi_gp2,
                         dt$hi_gp3,
                         dt$hi_gp4),
            ci.column = c(3, 5),
            ref.line = 1,
            ci.column.width = 2,
            ci.color = c("Group 1" = "#e41a1c", "Group 2" = "#4daf4a"),
            legend = list(name = "Group",
                          position = "bottom"),
            arrow.lab = c("Placebo Better", "Treatment Better"),
            nudge_y = 0.2)

# Draw plot
plot(p)


```

# TODO

- [ ] Expose more parameters, but keep it simple. Maybe treat it as a theme.
- [ ] More documentation
- [ ] Different point estimation symbols allow for different estimation type. 

